\chapterdated[A Letter to Max Shachtman]{A Letter to Max Shachtman\footnote{This letter was written by Trotsky in English.}}{December 20, 1939}
\label{1939-12-20_a-letter-to-max-shachtman}

\letteraddress{Dear Comrade Shachtman,}

I am sending you a copy of my last article.\footnote{The article referred to is ``A Petty-Bourgeois Opposition in the Socialist Workers Party.'' ---\ed} You will see from my polemics that I consider the divergences as of decisive character. I believe that you are on the wrong side of the barricades, my dear friend. By your position you give courage to all the petty-bourgeois and anti-Marxist elements to fight our doctrine, our program, and our tradition. I don’t hope to convince you with these lines, but I do express the prognosis that if you refuse now to find a way towards collaboration with the Marxist wing against the petty-bourgeois revisionists, you will inevitably deplore for years and years the greatest error of your life.

If I had the possibility I would immediately take an airplane to New York City in order to discuss with you for 48 or 72 hours uninterruptedly. I regret very much that you don’t feel in this situation the need to come here to discuss the questions with me. Or do you? I should be happy.

\signed{\textsc{L. Trotsky}}
\signed{Coyoacán, D.F.}

\begin{letterinfo}
	\firstpublished\ Leon Trotsky, \emph{In Defense of Marxism}, New York 1942.
	
	\checkedagainst\ Leon Trotsky, \emph{In Defence of Marxism}, London 1966, pp.~82.
\end{letterinfo}