\chapterdated[Letter to John G.\ Wright]{Letter to John G.\ Wright\footnote{This letter was written by Trotsky in English.}}{January 13, 1940}

\letteraddress{Dear Comrade Wright,}

I agree completely with your appreciation of the pamphlet of Comrade Shachtman.\footnote{The pamphlet referred to is ``The Crisis in the American Party---An Open Letter to Comrade Leon Trotsky,'' published in Vol.~II No.~7 of the \emph{Internal Bulletin}, dated January 1, 1940.} It is the \emph{weak} Shachtman multiplied by factional passion. He lacks a small thing which is called the proletarian point of view. He lives in the realm of literary shadows: when he stands with his face toward the proletariat and Marxism, his shadows are useful because they correspond more or less with reality; now he turns his back to the proletarian majority of the party and to Marxism and as a result every word he writes is a fantastic misinterpretation of facts and ideas. I am obliged now to lose again a couple of days to submit his absolutely extravagant document to a more attentive analysis. I hope to show to the party members including the majority of the minority that Shachtman’s document is in every line a pathetic rupture with Marxism and with Bolshevism.

\signed{Yours fraternally, \\ \textsc{L.\ Trotsky}}

\begin{letterinfo}
	\firstpublished\ Leon Trotsky, \emph{In Defense of Marxism}, New York 1942.
	
	\checkedagainst\ Leon Trotsky, \emph{In Defence of Marxism}, London 1966, p.~123.
	
	\footnoteslatter
\end{letterinfo}