% Copyright information
{
\setlength{\parindent}{0em}
\setlength{\parskip}{0.5em}
\sloppy

\vspace*{\fill}

Publisher’s Cataloging-in-Publication data

\vspace{0.5em}

{\setlength{\parskip}{0em}
Names: Morrow, Felix, 1906--1988, author.

Title: Revolution and Counter Revolution in Spain / Felix Morrow.

Description: Digital edition. | Madison, WI : 
attac.us, 2018.

Identifiers: LCCN 42039729 (print)

\hangindent=0.7cm
Subjects: Spain--Politics and government--1931--1939. Spain--History--Civil War, 1936-1939.

Classification: LCC DP269 .M615 1938a (print) | DDC 946.08

LC record available at \url{https://lccn.loc.gov/42039729}
}

\vspace*{\fill}

Originally published in 1938 by Pioneer Publishers, New York, NY.

Thanks to the Marxists Internet Archive
for making this book freely available.
Transcribed by Ted Crawford in 2003
and proofread by Einde O’Callaghan in August 2015.
Web version available at
\url{https://www.marxists.org/archive/morrow-felix/1938/revolution-spain/}

This edition designed by Jack Willis in 2018.
Typeset in 12pt Alegreya using the \LaTeX\ document preparation system.
Source code for this edition available at \url{https://www.attac.us/books/rcrs/}
}

\newpage

\chapter{Biographical notes}

This book was first published in 1938 by Pioneer in the United States.
An expanded edition which contained Morrow’s much shorter \emph{Civil War in Spain} pamphlet was published in the United States by Path\-finder Press in 1974 with a new index and copies may still be available from them. Note the reference to \emph{Civil War in Spain} by Morrow in Note 1 of Chapter 2, ``The Bourgeois ‘Allies’ in the People’s Front.''

The text of the following book is that of the second edition of 1963 published by New Park Publications together with the introduction written at that date by the late Tom Kemp. At the time, and for some years afterwards, the Socialist Labour League in Britain, who controlled New Park, was politically allied with the Socialist Workers Party in the United States, who controlled Pioneer/Pathfinder. The type of this edition was reset with English spellings etc. and the book was given its own index. Since it is now out of print, Index Books, the successor to New Park Publications, is consequently delighted that this important text is available, if only on the web, as they have no plans to reprint it.

It has been transcribed by Ted Crawford in 2003 and any errors and omissions are his responsibility. Notes have been placed at the end of the chapters and numbered rather than placed at the bottom of the page and the very occasional typographical errors removed. Any notes added are clearly stated to have been done by him and signed \emph{ERC}.

\begin{flushright}
Ted Crawford \\
2003
\end{flushright}