% Half-title
\begin{titlepage}
	\setlength{\parindent}{0pt}
	
	\vspace*{\fill}
	
	{\bfseries\sffamily\fontsize{42}{42}\selectfont
		Revolution\,\textnormal{and} \\
		Counter\,Revolution \\
		\textnormal{in}\,Spain
		\par}
	
	\vspace{32pt}
	
	\vspace*{\fill}
	
\end{titlepage}

% Full title
\begin{titlepage}
	\setlength{\parindent}{0pt}
	
	\vspace*{\fill}
	
	{\bfseries\sffamily\fontsize{42}{42}\selectfont
		Revolution\,\textnormal{and} \\
		Counter\,Revolution \\
		\textnormal{in}\,Spain
		\par}
	
	\vspace{11pt}
	
	{\normalfont\fontsize{21}{21}\selectfont
		\textsc{Felix Morrow}
		\par}
	
	\vspace*{\fill}
	
\end{titlepage}

% Copyright information
{
	\setlength{\parindent}{0em}
	\setlength{\parskip}{0.5em}
	\sloppy
	
	\vspace*{\fill}
	
	Publisher’s Cataloging-in-Publication data
	
	\vspace{0.5em}
	
	{\setlength{\parskip}{0em}
		Names: Morrow, Felix, 1906--1988, author.
		
		Title: Revolution and Counter Revolution in Spain / Felix Morrow.
		
		Description: Digital edition. | Madison, WI : 
		attac.us, 2018.
		
		Identifiers: LCCN 42039729 (print)
		
		\hangindent=0.7cm
		Subjects: Spain--Politics and government--1931--1939. Spain--History--Civil War, 1936-1939.
		
		Classification: LCC DP269 .M615 1938a (print) | DDC 946.08
		
		LC record available at \url{https://lccn.loc.gov/42039729}
	}
	
	\vspace*{\fill}
	
	Originally published in 1938 by Pioneer Publishers, New York, NY.
	
	Thanks to the Marxists Internet Archive
	for making this book freely available.
	Transcribed by Ted Crawford in 2003
	and proofread by Einde O’Callaghan in August 2015.
	Web version available at:
	\url{https://www.marxists.org/archive/morrow-felix/1938/revolution-spain/}
	
	This edition typeset in 12pt Alegreya by Jack Willis
	\texttt{<\href{mailto:jack@attac.us}{jack@attac.us}>}
	
	\textsc{pdf}s and \LaTeX\ source for this edition available at \url{https://www.attac.us/books/rcrs/}
	under a Creative Commons Attribution 4.0 International License:
	\url{https://creativecommons.org/licenses/by/4.0/}
}

% Space for dedication

\newpage
\thispagestyle{empty}
\ \\

\newpage
\thispagestyle{empty}
\ \\